---
output: github_document
---

```{r}
#| include: false
#| message: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
```

## About

This folder contains analytical datasets providing aggregated counts on the uptake of open access in hybrid journals included in transformative agreements by journal and country information between 2019 and 2023. A particular focus was on whether publications were from eligible authors who could use such agreements. For this purpose, the affiliations of  first and corresponding authors were analysed.

The following datasets were used:

- Scopus (`scp`): In-house Scopus database provided by the Kompetenznetzwerk Bibliometrie (April 2024 snapshot)
- Web of Science (`wos`): In-house Web of Science database provided by the Kompetenznetzwerk Bibliometrie (April 2024 snapshot)

See `data-raw/` how the subset of publications associated with hybrid journals and transformative agreements was retrieved. 
The indicator were determined using SQL. `indicators.R` contains the database queries.

## Data files

### Overview

Articles in hybrid journals by year
 
|Variable | Description |
|-------- |-----------  |
|`earliest_year`|Year first published |
|`pub_type`|`core`: Article or Review; `all`: All articles regardless of item type |
|`n_articles` | Number of items |
|`n_oa_articles` | Number of items in the database with a hybrid flag, indicating open access availability under an open license |
|`p`| Open access uptake |

#### `wos_by_year`

Web of Science

```{r}
#| message: false
readr::read_csv(here("data", "wos_by_year.csv"))
```

#### `scp_by_year`

Scopus

```{r}
#| message: false
readr::read_csv(here("data", "scp_by_year.csv"))
```

### By hybrid journal

Articles in hybrid journals by year and hybrid journal included in a transformative agreement

|Variable | Description |
|-------- |-----------  |
|`issn_l` | Linking ISSN, a unique identifier for different journal ISSN according to the ISSN agency |
|`earliest_year`|Year first published |
|`pub_type`|`core`: Article or Review; `all`: All articles regardless of item type |
|`n` | Number of items |
|`oa_articles` | Number of items in the database with a hybrid flag, indicating open access availability under an open license |

#### `wos_jn_by_year`

Web of Science

```{r}
#| message: false
readr::read_csv(here("data", "wos_jn_by_year.csv"))
```

#### `scp_jn_by_year`

Scopus

```{r}
#| message: false
readr::read_csv(here("data", "scp_jn_by_year.csv"))
```

### Country affiliations

Indicators are provided for both first and corresponding authors. Because of multiple country affiliations, full counting was applied. Note that the bibliometric database did not always include the corresponding affiliation.

|Variable | Description |
|-------- |-----------  |
|`issn_l` | Linking ISSN, a unique identifier for different journal ISSN according to the ISSN agency |
|`earliest_year`|Year first published |
|`country_code`|ISO three letter country codes |
|`pub_type`|`core`: Article or Review; `all`: All articles regardless of item type |
|`n` | Number of items |
|`oa_articles` | Number of items in the database with a hybrid flag, indicating open access availability under an open license |
|`first_author_articles`|Number of items from first authors|
|`oa_first_author_articles`|Number of open access items (tagged as hybrid) from first authors|
|`corresponding_author_articles`|Number of items from corresponding authors|
|`oa_first_author_articles`|Number of open access items (tagged as hybrid) from corresponding authors|
|`first_corresponding_author_articles`|Number of items from first authors, which are also corresponding authors|
|`oa_first_corresponding_author_articles`|Number of open access items (tagged as hybrid) from first authors, which are also corresponding authors|


#### `wos_country_aff.csv`

Web of Science 

```{r}
#| message: false
readr::read_csv(here("data", "wos_country_aff.csv"))
```

#### `scp_country_aff.csv`

Scopus

```{r}
#| message: false
readr::read_csv(here("data", "scp_country_aff.csv"))
```

### Transformative agreement data: Overview

The following data files contain indicators for journals where a link were established between articles and eligible institutions participating in transformative agreements. This could be achieved by linking author affiliations with agreement data.

Note that the earliest snapshot available with information about transformative agreemnt were from July 2021, while the analysis started in 2019.

|Variable | Description |
|-------- |-----------  |
|`earliest_year`|Year first published |
|`pub_type`|`core`: Article or Review; `all`: All articles regardless of item type |
|`ta_articles` | Number of items under an transformative agreement (first OR corresponding author)|
|`ta_oa_articles` | Number of items under an transformative agreement in the database with a hybrid flag, indicating open access availability under an open license (first OR corresponding author)|
|`ta_first_author_articles`|Number items from first authors under an transformative agreement |
|`ta_oa_first_author_articles`|Number of open access items (tagged as hybrid) from first authors under an transformative agreement|
|`ta_corresponding_author_articles`|Number of items from corresponding authors under an transformative agreement|
|`ta_oa_first_author_articles`|Number of open access items (tagged as hybrid) from corresponding authors under an transformative agreement|
|`ta_first_corresponding_author_articles`|Number of items from first authors under an transformative agreement, which are also corresponding authors|
|`ta_oa_first_corresponding_author_articles`|Number of open access items (tagged as hybrid) from first authors under an transformative agreement, which are also corresponding authors|

#### `wos_ta_by_year`

Web of Science

```{r}
#| message: false
readr::read_csv(here("data", "wos_ta_by_year.csv"))
```

#### `scp_ta_by_year`

Scopus

```{r}
#| message: false
readr::read_csv(here("data", "scp_ta_by_year.csv"))
```

### Transformative agreement data by hybrid journal

Articles in hybrid journals under transformative agreements by journal.
 
|Variable | Description |
|-------- |-----------  |
|`issn_l` | Linking ISSN, a unique identifier for different journal ISSN according to the ISSN agency |
|`earliest_year`|Year first published |
|`pub_type`|`core`: Article or Review; `all`: All articles regardless of item type |
|`ta_articles` | Number of items under an transformative agreement (first OR corresponding author)|
|`ta_oa_articles` | Number of items under an transformative agreement in the database with a hybrid flag, indicating open access availability under an open license (first OR corresponding author)|
|`ta_first_author_articles`|Number items from first authors under an transformative agreement |
|`ta_oa_first_author_articles`|Number of open access items (tagged as hybrid) from first authors under an transformative agreement|
|`ta_corresponding_author_articles`|Number of items from corresponding authors under an transformative agreement|
|`ta_oa_first_author_articles`|Number of open access items (tagged as hybrid) from corresponding authors under an transformative agreement|
|`ta_first_corresponding_author_articles`|Number of items from first authors under an transformative agreement, which are also corresponding authors|
|`ta_oa_first_corresponding_author_articles`|Number of open access items (tagged as hybrid) from first authors under an transformative agreement, which are also corresponding authors|

#### `wos_jn_ta_by_year`

Web of Science

```{r}
#| message: false
readr::read_csv(here("data", "wos_jn_ta_by_year.csv"))
```

#### `scp_ta_jn_by_year`

Scopus

```{r}
#| message: false
readr::read_csv(here("data", "scp_ta_jn_by_year.csv"))
```

### Transformative agreement data by country affiliations

Indicators are provided for both first and corresponding authors. Because of multiple country affiliations, full counting was applied. Note that the bibliometric database did not always include the corresponding affiliation.

|Variable | Description |
|-------- |-----------  |
|`issn_l` | Linking ISSN, a unique identifier for different journal ISSN according to the ISSN agency |
|`country_code`|ISO three letter country codes |
|`earliest_year`|Year first published |
|`pub_type`|`core`: Article or Review; `all`: All articles regardless of item type |
|`ta_articles` | Number of items under an transformative agreement (first OR corresponding author)|
|`ta_oa_articles` | Number of items under an transformative agreement in the database with a hybrid flag, indicating open access availability under an open license (first OR corresponding author)|
|`ta_first_author_articles`|Number items from first authors under an transformative agreement |
|`ta_oa_first_author_articles`|Number of open access items (tagged as hybrid) from first authors under an transformative agreement|
|`ta_corresponding_author_articles`|Number of items from corresponding authors under an transformative agreement|
|`ta_oa_first_author_articles`|Number of open access items (tagged as hybrid) from corresponding authors under an transformative agreement|
|`ta_first_corresponding_author_articles`|Number of items from first authors under an transformative agreement, which are also corresponding authors|
|`ta_oa_first_corresponding_author_articles`|Number of open access items (tagged as hybrid) from first authors under an transformative agreement, which are also corresponding authors|

#### `wos_jn_country_ta_by_year`

Web of Science

```{r}
#| message: false
readr::read_csv(here("data", "wos_jn_country_ta_by_year.csv"))
```

#### `wos_jn_country_ta_by_year`

Scopus

```{r}
#| message: false
readr::read_csv(here("data", "scp_jn_country_ta_by_year.csv"))
```

## hoadata
